Evaluating isolation and contact tracing for control of influenza outbreaks with pandemic potential: a modelling study

### 2025-07-02 inst/paper/main.tex
update paper .tex and .pdf and FluTracer.bib

\title{Evaluating isolation and contact tracing for control of influenza outbreaks with pandemic potential: a modelling study}
\author{Joshua W. Lambert, Authors TBC, Adam J. Kucharski}
\date{}

@@

Outbreaks of severe respiratory infections pose a major public health threat. The COVID-19 pandemic resulted in over 7 million reported deaths \citep{whocovid-19dashboardCOVID19DeathsWHO}, while the 1918 influenza pandemic caused an estimated $\sim$50 million deaths \citep{johnsonUpdatingAccountsGlobal2002}. 
Influenza subtypes to which populations have low levels of pre-existing immunity are a particular concern as potential pandemic candidates. This includes influenza A/H5N1 \citep{Ward2024.12.11.24318702}, A/H7N9 \citep{tannerPandemicPotentialAvian2015}, and A/H1N1p when it caused caused a pandemic in 2009 \citep{fraserPandemicPotentialStrain2009}. Recent human cases of avian influenza A/H5N1, alongside transmission, further illustrate the ongoing potential for viruses to infect new hosts, creating the potential for further adaption [CITE: Antia et al, Nature, 2003 + key papers on H5N1 - see my PLOS Bio piece]. It is therefore crucial to understand the effectiveness and feasibility of different intervention strategies in the face of such infections.

%Other pathogens, including those less transmissible, with longer delays between infections, and with less presymptomatic transmission, can also cause sustained outbreaks leading to the excess death of thousands \citep{whoebolaresponseteamEbolaVirusDisease2014}.  Timely and accurate response is therefore critical early in an outbreak when the pandemic or epidemic potential is still uncertain \citep{kucharskiControllingMinorOutbreaks2024}. An effective response requires understanding the efficacy and feasibility of intervention strategies across pathogens with varying epidemiological characteristics \citep{fraserFactorsThatMake2004}. \\ AK: Removed to improve flow, as this is mentioned in more detail later in intro.

%The likelihood of a novel outbreak causing sustained transmission depends both the characteristics of the pathogen and its host population, as well control measures  implemented in response to early cases. A key epidemiological metric to quantify the spread of an infectious disease is the reproduction number ($R$), defined as the average number of secondary cases by a typical infectious host. If the reproduction number is greater than 1, we would expect the incidence of infections to increase over time. If $R$ is below 1, we would expect incidence to decline on average, but several cases could still occur before transmission ceases \citep{farringtonDistributionTimeExtinction1999}. Two additional factors can influence the probability a pathogen with $R$ above 1 will successfully establish within a population~\citep{lloyd-smithSuperspreadingEffectIndividual2005}. If there is a large degree of individual-level variation in transmission, the majority of infectious individuals will infect nobody, even though a minority will infect many; as a result, most newly introduced outbreaks will fail to take off. In addition, the likelihood of establishment increases if there are multiple introductions into a populations \citep{whittleOutcomeStochasticEpidemicA1955, kucharskiEarlyDynamicsTransmission2020}. \\ % AK: Note 'primary case' means first in outbreak

Early in a novel outbreak, pharmaceutical options such as treatments and vaccines may not yet be available, leading to reliance on non-pharmaceutical interventions (NPIs) to reduce infection and disease. These interventions include individual-level targeted measures such as isolation and contact tracing, to population-level policies such as venue closures or restrictions on social mixing (Huag et al., 2020) [AK: probably a better reference out there, e.g. Sharma et al, Nature Comms]. The success of such interventions in curtailing transmission before it becomes sustained depends on several factors, including: the type of intervention, timing of introduction, and the intervention's effectiveness \citep{longiniContainingPandemicInfluenza2005}. For highly transmissible pathogens with a short generation times (e.g. pandemic influenza), incidence of new infections can grow quickly, outpacing a response. Moreover, pathogens such as influenza, which have a large proportion of onward transmission prior to symptom onset and have asymptomatic (subclinical) cases, are less easily controlled via isolation of symptomatic cases and tracing and quarantining of their contacts~\citep{fraserFactorsThatMake2004}.

Several pre-COVID-19 studies concluded that influenza viruses cannot be controlled by isolating individuals at symptom onset and tracing their contacts, because the rate at which infected individuals become infectious and transmit is faster than the latency in the process of isolating contacts \citep{fraserFactorsThatMake2004, klinkenbergEffectivenessContactTracing2006}. However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers \citep{gargHighlyPathogenicAvian2025} and the longer estimates of incubation period and serial interval from historical H5N1 data \citep{Ward2024.12.11.24318702} it may be that isolation and contact tracing is able to control influenza subtypes with pandemic potential. \\

%Isolation of symptomatic infections aims to break the chain of transmission by preventing further infection of susceptible individuals. The earlier an infectious case can be isolated, the more effective it is at averting secondary cases, with isolation prior to becoming infectious optimal. This can make it challenging to control outbreaks with symptom-based isolation, as for most pathogens infectiousness preceeds prodromal symptoms [AK: reference? This is a strong claim – best to focus on flu I think]. Quarantining contacts exposed to an infectious individuals, independent of whether they later become a case, is a control strategy to reduce transmission before or shortly after symptom onset. Contact tracing is the process of finding the contacts of infected individuals. Both isolation and contact tracing are often used NPIs to try and curtail the growth of an epidemic. Forward contact tracing asks infected individuals who they have been in contact with so their contacts can be followed-up and isolated or quarantined if needed. Isolation, quarantine, and contact tracing is a targeted approach that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts \citep{ferrettiQuantifyingSARSCoV2Transmission2020, keelingEfficacyContactTracing2020}. Isolation or vaccination of traced individuals has proved widely effective in controlling different outbreaks with a variety of pathogens differing in their aetiological and epidemiological characteristics \citep{foegeSELECTIVEEPIDEMIOLOGICCONTROL1971, bellPublicHealthInterventions2004}. However, in epidemics or pandemics where the proportion of cases or contacts missed is high, either because surveillance is patchy or because the number of contacts exceeds the capacity of the system, the efficacy of contact tracing to contain an outbreak diminishes, or at worst, fails completely \citep{dhillonWhenContactTracing2018, hellewellFeasibilityControllingCOVID192020}. \\ AK: This is probably too much fundamental background for an applied journal – readers will know what isolation and contact tracing is.

%The (antigenic) novelty of a pathogen introduced into a human population is a leading risk factor for causing a pandemic. Thus, Historically, avian influenza (H5N1 and H7N9) has had high severity (case fatality risk $\sim 35\%-55\%$), and therefore sustained transmission in a human population could have substantial mortality and morbidity implications \citep{tannerPandemicPotentialAvian2015}. \\ [AK: These headline fatality risks are likely to be biased: https://kucharski.substack.com/p/how-fatal-is-h5n1-influenza]

In this study we evaluate the effectiveness of controlling outbreaks that exhibit influenza-like epidemiological characteristics using a mathematical model of isolation and contact tracing. In particular, we investigate the assumption that influenza will inevitably outpace ability to trace and isolate sufficiently effectively for control, and whether this assumption holds for all pathogen subtypes given emerging estimates of incubation period and serial interval \citep{Ward2024.12.11.24318702}. Finally, we quantify the speed and effectiveness of isolation for an effective response under different response delays across epidemic scenarios, and assess the impact of reducing the delay to isolation on outbreak containment alongside use of rapid testing.

@@

\subsection*{Transmission model}
We used a branching process transmission model to assess the potential effectiveness of isolation and contact tracing to control an outbreak of pandemic-potential influenza \texttt{\{ringbp\}} \citep{hellewellRingbpSimulateEvaluate2025} (Figure \ref{fig:ringbp-model}). In the model, each infectious individual generates secondary cases from a negative binomial distribution with a given reproduction number $R$ and dispersion parameter ($k$). Onwards transmission from infectious individuals could be reduced in the following ways: 1) \textit{symptomatic individuals isolate effectively}, 2) \textit{contacts of symptomatic individuals are traced and isolate if they become symptomatic}, 3) \textit{contacts of symptomatic individuals are traced and effectively quarantined immediately}. Symptomatic individuals that have been traced (i.e. infectees) are isolated either 1) after the onset-to-isolation delay after symptom onset, or 2) when their infector is isolated. The isolation time of the infectee is whichever occurs first of these two events.

We include several mechanisms by which isolation and contact tracing can be imperfect at reducing onward disease transmission. The proportion of contacts identified by contact tracing can range between 0\% (i.e. no contacts are traced) to 100\% (i.e. every contact of every symptomatic case is traced). Because the model focuses on onwards transmission prevented, identifying 50\% of cases and preventing 100\% of their onwards transmission via contact tracing is equivalent to identifying 100\% of cases and preventing 50\% of their onwards transmission. Subsequent symptomatic cases that have not been contact traced are only isolated after the onset-to-isolation delay post symptom onset. Additionally, we assume infections have an independent probability of being asymptomatic or paucisymptomatic; these infections were not identified and therefore their contacts were not traced. The advantage of this approach is that it enables estimation of the overall population transmission effect of multiple interventions without requiring each step to be individually specified as a parameter.

\subsection*{Pathogen epidemiological parameters} \label{epiparameters}

In this study we evaluate the feasibility of controlling outbreaks of A/H5N1, A/H1N1, and A/H7N9-like infections.\cite{nishiuraEstimationIncubationPeriod2011} report the best fitting Weibull shape and scale parameters for H1N1 to be 1.77 and 1.86, respectively (Figure \ref{fig:delay-distributions}A). \cite{cowlingComparativeEpidemiologyHuman2013} report the incubation period for H5N1 and H7N9, both Weibull distributed. For H5N1 the Weibull distribution has a mean of 3.3 days and standard deviation (SD) of 1.5 days, for H7N9 the Weibull distribution has a mean of 3.1 days and SD 1.4 days \citep{cowlingComparativeEpidemiologyHuman2013} (Figure \ref{fig:delay-distributions}A). Comparatively, the incubation period for H1N1 shorter than the other two subtypes (H5N1 and H7N9), which are similar, with H5N1 having a slightly longer median (Figure \ref{fig:delay-distributions}A).  We consider four reproduction number ($R$) values when simulating transmission: 1.1, 1.5, 2.5 and 3.5. These are based on similar values to previously estimated $R$ values for influenza outbreaks \citep{fergusonStrategiesMitigatingInfluenza2006}, including H1N1 \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}, or they represent inflated values of current estimates of H5N1 and H7N9 subtypes to evaluate the effectiveness of isolation and contact tracing in the event that pathogen evolution.

We three presymptomatic transmission scenarios: <1\%, 15\%, 30\% of transmission occurring prior to symptom onset. % AK: Probably need a 50%?
We assumed that asymptomatic infections can occur and infect others, but are less common that symptomatic infectious indivdiuals and that the majority of cases will develop symptoms during their infectious period.  We ran three scenarios, one with zero asymptomatic illness, another with 10\% of cases being asymptomatic and the most extreme scenario where 30\% of cases never develop symptoms and thus are not isolated and secondary cases are missed by contact tracing.

@@

We define outbreak control as no new cases between 12 and 16 weeks after the start of an outbreak\cite{hellewellFeasibilityControllingCOVID192020}. The rationale for this criteria is that some outbreaks that do not take off and become large may still have a few cases for the first few weeks of an outbreak, therefore the window for control needs to be enough time after the seeding of an outbreak to be sure it is extinct. There is also a maximum number of cases in simulation model, that when reached is judged to have reached a large epidemic and not controllable within 12-16 weeks. We set this limit to 500 cases. To test scenarios where an outbreak starts with different number of initial infections, we simulate with either 5, 20 or 40 initial independent cases. This could represent importation of multiple cases in a single event, such as a flight with multiple positive cases. \\

The onset-to-isolation delay is parameterised with three scenarios: a fast response, based on the SARS in Hong Kong (Donnelly et al. 2003) and a slower response based on COVID-19 in Wuhan, China (Li et al. 2020) (Figure \ref{fig:delay-distributions}B). Both the fast and slow onset-to-isolation are parameterised as Weibull distributions, the SARS-like distribution has shape ($\lambda$) 1.65 and scale ($k$) 4.29 (mean = 3.83 days, SD = 2.38 days), the COVID-like distribution has shape = 2.31 and scale = 9.48 (mean = 8.40, SD = 3.87 days) (Figure \ref{fig:delay-distributions}B). We also include a third onset-to-isolation scenario based on the case where lateral flow rapid tests (LFTs) are provided to all individuals in the susceptible population and these people are freely able to self-test if they suspect they are symptomatic. We parameterise this as an exponential distribution, with rate ($\lambda$) 0.5, where most individuals will self-test very shortly after they become symptomatic (approximately 40\% of cases will self-test within the first day of sympmtom onset), with a few people waiting a 2 or more days before testing (approximately 3\% take longer than one week to take a self-test). We assume that LFTs are 100\% accurate so all symptomatic individuals will receive a positive test result.

@@

The ability to control an influenza epidemic is highly dependent on the reproduction number and the speed of isolation after symptom onset (Figure \ref{fig:prop-outbreak-control-R}). A higher outbreak reproduction number results in a smaller percentage of outbreaks controlled at a given percentage of contacts traced, except in cases where they are equal because zero or all outbreaks are contained, across all onset-to-isolation scenarios (Figure \ref{fig:prop-outbreak-control-R}). When the reponse to isolate cases is slow then outbreaks with $R = 1.5$ require at least 50\% of contacts traced to contain at least half of simulated outbreaks. When $R$ is 2.5 or 3.5 then contact tracing needs to be extremely comprehensive ($>90\%$) to have a chance at containment, and even then $R = 3.5$ influenza outbreaks on average are uncontrolled (Figure \ref{fig:prop-outbreak-control-R}A). Especially for H1N1, given its faster tempo of transmission, has a much lower probability of containment at 100\% of contacts traced with a slow, COVID-19-like, isolation response time (Figure \ref{fig:prop-outbreak-control-R}A). In a scenario with a fast, Hong Kong SARS-like, onset-to-isolation response, outbreaks with $R = 1.1$ are almost always contained at any proportion of contacts traced (Figure \ref{fig:prop-outbreak-control-R}B). For slow and fast isolation response delays there is a substantial difference between outbreaks with a $R$ slightly above unity ($R = 1.1$) which were controlled at a low levels ($\geq 20\%$) of isolation and contact tracing for all three influenza subtypes (Figure \ref{fig:prop-outbreak-control-R}A,B). As $R$ increases to 1.5 there is a dramatic decline in the ability to control the epidemic without isolation of traced contacts (i.e. 0\% contact ascertainment), however, outbreaks with $R = 1.5$ are mostly contained once the percentage of contacts ascertained in tracing exceeds 50\% (Figure \ref{fig:prop-outbreak-control-R}B). For more transmissible influenza scenarios, even a fast isolation response, with an average of 3.4 days between symptom onset and isolation, requires $>70\%$ of contacts traced for most outbreaks to be controlled (Figure \ref{fig:prop-outbreak-control-R}B). \\

The most rapid scenario we simulated to emulate individuals self-testing with LFTs, has the same trend of increased chance of outbreak containment by reducing the time between symptom onset and isolation, as the slow and fast scenarios (Figure \ref{fig:prop-outbreak-control-R}). The control of low transmissible ($R \in {1.1, 1.5})$ influenza pathogens is almost always guranteed even in the absence of contact tracing and only isolating cases immediately if positive on an LFT (Figure \ref{fig:prop-outbreak-control-R}C). Assuming ubiquitous access to home testing and that 63\% of symptomatic cases test and isolate within the 24 hours symptom onset, 75\% ascertainment of contacts results in most outbreaks controlled for all influenza transmissibility scenarios ($R \leq 3.5$). The LFT isolation strategy is also the only onset-to-isolation delay which can achieve 100\% of outbreaks controlled when all contacts are traced (Figure \ref{fig:prop-outbreak-control-R}C). \\

The difference in influenza subtype incubation period influences the efficacy of isolating and tracing. This effect is most pronounced when the percentage of outbreaks controlled is greater than 0\% and less than 100\% (i.e. not at the boundary) (Figure \ref{fig:prop-outbreak-control-R}). The shortest incubation period, H1N1, is predominantly the least controllable for a given reproduction number, showing that infections that are symptomatic shortly after infection and in some instances transmission while subclinical limit our ability to effectively trace and isolate their potential infectees in time to prevent disease transmission. The variance between the influenza subtypes at each percentage of contacts traced and isolated varies (i.e. difference between circle, triangle and square points at each contacts traced \% in Figure \ref{fig:prop-outbreak-control-R}). This variance is greatest at intermediate (20\% - 80\%) contact tracing effectiveness and intermediate values of $R$ (1.5 and 2.5) (Figure \ref{fig:prop-outbreak-control-R}). This could indicate a substantive role for incubation period durartion in the control of disease transmission. However, this variance in proportion of outbreaks controlled between pathogen subtypes does not show a strong pattern across all scenarios simulated (Figure \ref{fig:prop-outbreak-control-var-R}). There is a weak pattern of increased variance in controllability between subtypes at intermediate values of $R$, overall it seems that incubation period causes differences in control-by-isolation effectivness when not at the boundary of epidemic controllability. In other words, differences between influenza subtypes arise when not in a situation of high transmissibility and low contact tracing effectivness that cannot contain any outbreaks, nor low transmissibility and high contact tracing effectivenss easily contain all outbreaks (Figure \ref{fig:prop-outbreak-control-var-R}). \\

@@

>>>>

\title{Evaluating isolation and contact tracing for control of influenza outbreaks with pandemic potential: a modelling study}
\author{Joshua W. Lambert, Adam J. Kucharski}
\date{}

@@

Outbreaks of severe respiratory infections pose a major public health threat. The COVID-19 pandemic resulted in over 7 million reported deaths \citep{whocovid-19dashboardCOVID19DeathsWHO}, while the 1918 influenza pandemic caused an estimated $\sim$50 million deaths \citep{johnsonUpdatingAccountsGlobal2002}. Influenza subtypes to which populations have low levels of pre-existing immunity are a particular concern as potential pandemic candidates. This includes influenza A/H5N1 \citep{Ward2024.12.11.24318702} and A/H7N9 \citep{tannerPandemicPotentialAvian2015}, as well as A/H1N1pdm when it caused a pandemic in 2009 \citep{fraserPandemicPotentialStrain2009}. Recent human cases of avian influenza A/H5N1, alongside transmission, further illustrate the ongoing potential for viruses to infect new hosts, creating the potential for further adaption \citep{antiaRoleEvolutionEmergence2003, gargHighlyPathogenicAvian2025, peacockGlobalH5N1Influenza2025}. It is therefore crucial to understand the effectiveness and feasibility of different intervention strategies in the face of such infections \citep{fraserFactorsThatMake2004, kucharskiControllingMinorOutbreaks2024}. \\

Early in a novel outbreak, pharmaceutical options such as treatments and vaccines may not yet be available, leading to reliance on non-pharmaceutical interventions (NPIs) to reduce infection and disease. These interventions include individual-level targeted measures such as isolation and contact tracing, to population-level policies such as venue closures or restrictions on social mixing \citep{haugRankingEffectivenessWorldwide2020, sharmaUnderstandingEffectivenessGovernment2021}. The success of such interventions in curtailing transmission before it becomes sustained depends on several factors, including: the type of intervention, timing of introduction, and the intervention's effectiveness \citep{longiniContainingPandemicInfluenza2005}. For highly transmissible pathogens with a short generation times (e.g. pandemic influenza), incidence of new infections can grow quickly, outpacing a response. Moreover, pathogens such as influenza, which have a large proportion of onward transmission prior to symptom onset and have asymptomatic (subclinical) cases, are less easily controlled via isolation of symptomatic cases and tracing and quarantining of their contacts \citep{fraserFactorsThatMake2004}. \\

Several pre-COVID-19 studies concluded that influenza viruses cannot be controlled by isolating individuals at symptom onset and tracing their contacts, because the rate at which infected individuals become infectious and transmit is faster than the latency in the process of isolating contacts \citep{fraserFactorsThatMake2004, klinkenbergEffectivenessContactTracing2006}. However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers \citep{gargHighlyPathogenicAvian2025} and the longer estimates of incubation period and serial interval from historical A/H5N1 data \citep{Ward2024.12.11.24318702} it may be that isolation and contact tracing is able to control influenza subtypes with pandemic potential. \\

In this study we evaluate the effectiveness of controlling outbreaks that exhibit influenza-like epidemiological characteristics using a mathematical model of isolation and contact tracing. In particular, we investigate the assumption that influenza will inevitably outpace our ability to trace and isolate sufficiently effectively for control, and whether this assumption holds for all pathogen subtypes given emerging estimates of incubation period and serial interval \citep{Ward2024.12.11.24318702}. Finally, we quantify the speed and effectiveness of isolation for containment under different response delays across epidemic scenarios, and assess the impact of reducing the delay to isolation on outbreak control alongside the use of rapid testing.

@@

\subsection*{Transmission model}

We used a branching process transmission model to assess the potential effectiveness of isolation and contact tracing to control an outbreak of pandemic-potential influenza, implemented in the R package \texttt{\{ringbp\}} (Figure \ref{fig:ringbp-model}) \citep{hellewellRingbpSimulateEvaluate2025}. In the model, each infectious individual generates secondary cases from a negative binomial distribution with a given reproduction number ($R$) and dispersion parameter ($k$). Onwards transmission from infectious individuals could be reduced in the following ways: 1) symptomatic individuals isolate effectively, 2) contacts of symptomatic individuals are traced and isolate if they become symptomatic, 3) contacts of symptomatic individuals are traced and effectively quarantined immediately. Symptomatic individuals that have been traced (i.e. infectees) are isolated at whichever of two events occurs first, either 1) after the onset-to-isolation delay after symptom onset, or 2) when their infector is isolated (if quarantined or symptomatic) or when they become symptomatic if the infector is already isolating. \\

We include several mechanisms by which isolation and contact tracing can be imperfect at reducing onward disease transmission. The proportion of contacts ascertained by contact tracing can range between 0\% (i.e. no contacts are traced) to 100\% (i.e. every contact of every symptomatic case is traced). Because the model focuses on onwards transmission prevented, identifying 50\% of cases and preventing 100\% of their onwards transmission via contact tracing is equivalent to identifying 100\% of cases and preventing 50\% of their onwards transmission. Subsequent symptomatic cases that have not been contact traced are only isolated after the onset-to-isolation delay post symptom onset. Additionally, we assume infections have an independent probability of being asymptomatic (or paucisymptomatic); these infections were not identified and therefore their contacts were not traced. The advantage of this approach is that it enables estimation of the overall population transmission effect of multiple interventions without requiring each step to be individually specified as a parameter.

\begin{figure}[ht]
\centering
\includegraphics[width=\textwidth]{../plots/model-schematic.png}
\caption{Model schematic of the branching process with isolation, quarantine and contact tracing implemented in the \texttt{\{ringbp\}} R package. (A) The transmission chain from the index case (Person A), infection events are shown by black vertical dashed lines. Individuals can be infected (red circle), become symptomatic (orange circle), and isolated (blue circle). Quarantine is optional in the model, assuming quarantine is active contacts are quarantined when the infector is isolated independent of the contacts symptom status (green circle with Q). The time at which a infectors isolation triggers the isolation of a symtomatic case is shown by the blue vertical dashed line (Person A to Person B). The isolation times that would have been triggered after a delay from symptom onset, but are after isolation was triggered by contact tracing is shown by a blue circle with a cross and connected by a dotted line (Person B). Person C is an asymptomatic individual (asymptomatic cases can infect others). Person D is missed by contact tracing (contacts are traced with a probability $\rho$, and thus missed at $1 - \rho$). The time Person D would have isolated if ascertained is shown by the vertical blue mark. Assuming quarantine is active, Person F would be isolated before symptom onset at the time of the infector's (Person D) isolation time. If quarantine is not active, then Person F would be isolated after a delay from symptom onset, as they had yet to show symptoms when the infector was isolated. (B) A branching process diagram showing who-infected-whom.}
\label{fig:ringbp-model}
\end{figure}

\subsection*{Pathogen epidemiological parameters}

In this study we evaluate the feasibility of controlling outbreaks of A/H5N1-, A/H1N1-, and A/H7N9-like infections. The generation time is modelled using the incubation period and the proportion of presymptomatic transmission to simulate a correlated structure between the two delay distributions to generate sensible delays between exposure, symptom onset and onward transmission (Figure \ref{fig:delay-distributions}A) \citep{hellewellFeasibilityControllingCOVID192020, lintonCorrelationTimesSARSCoV22022}. We extracted previously estimated incubation periods from the literature for each subtype. \cite{nishiuraEstimationIncubationPeriod2011} report the best fitting Weibull shape and scale parameters for H1N1 to be 1.77 and 1.86, respectively (Figure \ref{fig:delay-distributions}B). \cite{cowlingComparativeEpidemiologyHuman2013} report the incubation period for H5N1 and H7N9, both Weibull distributed. For H5N1 the Weibull distribution has a mean of 3.3 days and standard deviation (SD) of 1.5 days, for H7N9 the Weibull distribution has a mean of 3.1 days and SD 1.4 days \citep{cowlingComparativeEpidemiologyHuman2013} (Figure \ref{fig:delay-distributions}B). Comparatively, the incubation period for H1N1 shorter than the other two subtypes (H5N1 and H7N9), which are similar, with H5N1 having a slightly longer median (Figure \ref{fig:delay-distributions}B). \\

We simulated three presymptomatic transmission scenarios: 1\%, 15\%, and 30\% of transmission occurring prior to symptom onset. We assumed that asymptomatic infections can occur and infect others, but are less common that symptomatic infectious indivdiuals and that the majority of cases will develop symptoms during their infectious period.  We ran scenarios with three different assumed proportion of asymptomatic cases, one with zero asymptomatic illness, another with 10\% of cases being asymptomatic and the most extreme scenario where 30\% of cases never develop symptoms and thus are not isolated and secondary cases are missed by contact tracing. These proportions of presymptomatic transmission and asymptomatic fraction cover a broad range of possible pandemic influenza possibilities (Figure \ref{fig:patho-param-space}) \citep{lesslerOutbreak2009Pandemic2009, xuSerologicalInvestigationSubclinical2013, montgomeryRoleAsymptomaticInfections2024, gargHighlyPathogenicAvian2025}. \\

\begin{figure}[ht]
\centering
\includegraphics[width=\textwidth,height=0.85\textheight]{../plots/delay_distributions.png}
\caption{(A) Proportion of presymptomatic transmission determined by the incubation period, here assumed to be 5 days (dashed vertical line), controlled by a skew-normal distribution. Three skew-normal distribution probability density functions are plotted with different skew parameters, 0.73, 1.96, 31.82, which correspond to 30\%, 15\% and 1\% presymptomatic transmission. (B) Incubation period for the influenza subtypes: A/H1N1, A/H5N1 and A/H7N9. The vertical dashed lines are the median incubation period: A/H1N1 is 1.5 days (yellow dashed line), A/H5N1 is 3.2 days (purple dashed line), and A/H7N9 is 3.0 days (blue dashed line). (C) Three assumed onset-to-isolation delays used in the simuluation experiment. The \textit{Slow} scenario is based on the onset-to-isolation for the SARS-CoV-2 response in Wuhan, China \citep{liEarlyTransmissionDynamics2020}, the median onset-to-isolation is 8.1 days (blue dashed line). The \textit{Fast} scenario is based on the onset-to-isolation for SARS-CoV-1 response in Hong Kong \citep{donnellyEpidemiologicalDeterminantsSpread2003a}, the median onset-to-isolation time is 3.4 days (red dashed line). The lateral flow test (LFT) scenario is an assumed distributed for rapid self-administered home testing upon isolation, the median delay from symptom onset to isolation in this scenario is 1.4 days (orange dashed line).}
\label{fig:delay-distributions}
\end{figure}

We considered four reproduction number ($R$) values when simulating transmission in the community: 1.1, 1.5, 2.5 and 3.5 (Figure \ref{fig:patho-param-space}). These are based on similar values to previously estimated $R$ values for influenza outbreaks \citep{fergusonStrategiesMitigatingInfluenza2006}, including A/H1N1pdm \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}, or they represent inflated values of current estimates of A/H5N1 and A/H7N9 subtypes to evaluate the effectiveness of isolation and contact tracing in the event that pathogen evolution enhances the human-to-human transmissibility (e.g. genetic reassortment \citep{peacockGlobalH5N1Influenza2025} or mutations in receptor binding preference \citep{linSingleMutationBovine2024}). Once isolated, cases are assumed to not infect anyone ($R_{iso} = 0$). The reproduction number of asymptomatic cases is assumed equal to the community. The dispersion paramter ($k$) for the negative binomial offspring distributions was fixed at 0.8 for community and asymptomatic cases following evidence that influenza subtypes have moderate transmission heterogeneity \citep{fraserPandemicPotentialStrain2009, heComparingCOVID191918192020, Ward2024.12.11.24318702}.

\subsection*{Scenarios}

The onset-to-isolation delay is parameterised under three response scenarios. A fast response, based on the SARS in Hong Kong \citep{donnellyEpidemiologicalDeterminantsSpread2003a} and a slower response based on COVID-19 in Wuhan, China \citep{liEarlyTransmissionDynamics2020} (Figure \ref{fig:delay-distributions}C). Both the fast and slow onset-to-isolation are parameterised as Weibull distributions, the SARS-like distribution has shape ($\lambda$) 1.65 and scale ($k$) 4.29 (mean = 3.83 days, SD = 2.38 days), the COVID-like distribution has shape = 2.31 and scale = 9.48 (mean = 8.40, SD = 3.87 days) (Figure \ref{fig:delay-distributions}C). The third onset-to-isolation scenario is based on a situation where lateral flow rapid tests (LFTs) are provided to all susceptible individuals in the population and these people are freely able to self-test if they are symptomatic. We parameterise this as an exponential distribution, with rate ($\lambda$) 0.5, where most individuals will self-test very shortly after they become symptomatic (approximately 40\% of cases will self-test within the first day of sympmtom onset), with a few people waiting a 2 or more days before testing (approximately 3\% take longer than one week to take a self-test) (Figure \ref{fig:delay-distributions}C). We assume that LFT sensitivity is 100\% accurate so all symptomatic individuals will receive a positive test result. \\

To assess the feasibility of outbreak containment under different proportions of contacts traced and isolated we varied the proportion of contacts ascertained between 0\% (i.e. no contacts traced) and 100\% (i.e. all contacts traced) at intervals of 20\%. When contact ascertainment is zero, isolation is the only NPI reducing transmission. To test scenarios where an outbreak starts with different number of initial infections, we simulate with either 5, 20 or 40 initial independent cases. This could represent importation of multiple cases in a single event, such as a flight. Lastly, we determine if addtionally quarantining contacts, that is isolating contacts once the infector is isoloted independent of whether the contact is symptomatic, improves the ability to contain influenza outbreaks.

@@

We define outbreak control as no new cases after 12 weeks from the start of an outbreak. The rationale for this criteria is that some outbreaks that do not take off and become sustained may still have cases for the first few weeks of an outbreak, therefore the window for control needs to be enough time after the seeding of an outbreak to be sure it is contained \citep{hellewellFeasibilityControllingCOVID192020}. Due to the exponential dynamics of outbreaks and no depletion of susceptibles in our model we implement a maximum number of cases in each simulation. When this cumulative case limit is reached, it is judged to be a sustained epidemic, not controllable within 12 weeks. We set this limit to 5,000 cases.

@@

The reproduction number, contact tracing ascertainment and speed of isolation all affect the ability to control a pandemic potential influenza outbreak (Figure \ref{fig:prop-outbreak-control-R}). Isolation alone can control over 50\% of epidemics with an $R$ of 1.1 with a slow onset-to-isolation response (median 8.1 days); while a fast (median 3.4 days) or LFT (median 1.4 days) response almost always controls $R = 1.1$ outbreaks (Figure \ref{fig:prop-outbreak-control-R}). When $R = 1.5$, $>$75\% of contacts need to be traced to control at least 75\% of outbreaks for a slow onset-to-isolation, whereas only 40\% of contacts require tracing for at least 75\% controlled under a fast isolation delay (Figure \ref{fig:prop-outbreak-control-R}A,B). The isolation upon rapid LFT self-testing controls epidemics with $R \leq 1.5$ without tracing contacts (Figure \ref{fig:prop-outbreak-control-R}C). High transmissibility scenarios ($R \geq 2.5$) require comprehensive contact tracing to contain at least 50\% of simulated outbreaks ($\sim$ 100\% for a slow response and $\sim$ 80-100\% for a fast response). Isolation and tracing 80\% of contacts using LFTs to trigger isolation controls over 80\% of simulated outbreaks for all subtype and transmissibility scenarios (Figure \ref{fig:prop-outbreak-control-R}C). \\

@@

There is some variability in the effectiveness of isolation and contact tracing for outbreak control between influenza subtypes. A/H1N1, because of its shorter incubation period and thus faster tempo of transmission, has a lower proportion of outbreaks controlled, most noticably for $R \geq 1.5$ (Figure \ref{fig:prop-outbreak-control-R}). There is not a strong pattern for the variance in outbreak control between influenza subtypes across scenarios; the few scenarios with higher variance between subtypes are at high percentages of contacts traced ($\geq$ 60\%) and for high transmissibility scenarios (Figure \ref{fig:prop-outbreak-control-var-R}). \\

Increased presymptomatic transmission and asymptomatic cases decreases epidemic containment (Figure \ref{fig:prop-outbreak-control-prop-presym-iso} and \ref{fig:prop-outbreak-control-prop-asym-iso}). When 30\% of transmission occurs before symptom onset, in high transmissibility scenarios ($R \geq 2.5$) more than half of outbreaks become sustained epidemics for all onset-to-isolation delays (Figure \ref{fig:prop-outbreak-control-prop-presym-iso}). When only 1\% of transmission is presymptomatic and contact tracing ascertains all contacts, outbreaks with $R = 3.5$ are contained in $>$75\% of outbreaks with LFT testing, and in $\sim$25\% of outbreaks with a fast isolation response (Figure \ref{fig:prop-outbreak-control-prop-presym-iso}). When all cases develop symptoms, isolation and contact tracing all contacts can control all outbreak scenarios, with the sole exception being A/H1N1 with $R = 3.5$ and a slow onset-to-isolation (Figure \ref{fig:prop-outbreak-control-prop-asym-iso}). When 10\% of cases are asymptomatic the ability to contain highly transmissible influenza ($R \geq 2.5$) dramatically decreases, and decreases further when 30\% of cases are asymptomatic or paucisymptomatic, even with comprehensive contact tracing and isolation upon LFT self-testing, less than a quarter of outbreaks are contained (Figure \ref{fig:prop-outbreak-control-prop-asym-iso}). \\

The more cases that initially seed an outbreak with $R \geq 1$, the less chance that it will stochastically go extinct and we find that the higher the number of initial influenza cases reduces the efficacy of control-by-isolation (Figure \ref{fig:prop-outbreak-control-num-init-cases}). Isolation alone can control outbreaks with $R = 1.1$ even when seeded with 40 initial cases (Figure \ref{fig:prop-outbreak-control-num-init-cases}). When $R \geq 1.5$ increasing the number of initial cases makes epidemics less controllable for all influenza subtypes. For high transmissibility ($R \geq 2.5$) scenarios that are seeded with at least 20 cases, close to 100\% of contacts need to be traced and isolated upon symptom onset to control the majority of outbreaks, assuming a fast onset-to-isolation response (3.4 days on average) (Figure \ref{fig:prop-outbreak-control-num-init-cases}). \\

In the acute phase of outbreaks there can be thousands of weekly cases, especially for high transmissibility scenarios ($R \geq 2.5$) (Figure \ref{fig:max-weekly-cases}). Specifically for controlled outbreaks, the cumulative number of cases is consistently below 500 cases across influenza subtypes (Figure \ref{fig:median-controlled-outbreak-size}). Thus when an outbreak is controlled with isolation alone or in combination with contact tracing it is done so without incurring a high initial prevalence. \\
